---
title: "LIFE748 - ML and data analysis"
format: html
editor: visual
---

## Introduction

In this tutorial, we will analyze a simulated RNA-seq dataset. Our goal is to identify Differentially Expressed Genes (DEGs) between two conditions (Condition A and Condition B), visualize the results, and use both unsupervised and supervised machine learning techniques to classify the samples based on their gene expression profiles.

## Declaration

AI Declaration Claude (Anthropic, claude.ai) was used to assist with the following: - Interpreting the confusion matrix outputs from LDA and SVM classifiers - Understanding the difference between single split and cross-validation accuracy - Identifying appropriate figure types for visualising classifier performance.

## Prompts used:

why did all condition b samples cluster together but condition a split - my lda cv accuracy ranges from 0.40 to 0.90 why so variable w - what does a kappa of 0.57 mean is that good for svm - how do i interpret sensitivity vs specificity for my classifiers - why does svm out perform lda on gene expression data - how do i run 5 fold cross validation for lda and svm in r - how do i use groupdata2 to balance folds by condition - my confusionmatrix function is not recognising factor levels how fix - tthe way to i extract accuracy sensitivity and specificity from confusionmatrix output ? - Please check my code for th bar chart comparing lda and svm performance metrics in ggplot2

All analysis was executed by me using the provided QMD template.

## Step 0: Load the required libraries

```{r warning=FALSE, message=FALSE }
library(DESeq2) 
library(ggplot2) 
library(EnhancedVolcano) 
library(pheatmap) 
library(caret)
library(MASS) 
library(e1071) 
library(groupdata2)
```

If these packages are not already available, you will need to install them either from CRAN (install.packages('package_name')), or from Bioconductor (BiocManager::install('package_name')

## Step 1: Load the Count Matrix

First, we load the raw count matrix and set up the sample metadata (phenotype data). We assume the first 25 columns belong to Condition A, and the next 25 belong to Condition B.

```{r}
counts <- read.csv("dataset_LIFE748_3.csv", row.names = 1)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = c(rep('CondA',25),rep('CondB',25))
)
```

## Step 2: Normalize and Identify DEGs with DESeq2

We will use `DESeq2` to normalize the library sizes and perform statistical testing to identify which genes are differentially expressed between Condition B and Condition A.

```{r}
# 1. Create the DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = colData,
                              design = ~ Condition)

# 2. Run the DESeq2 pipeline
dds <- DESeq(dds)

# 3. Extract the results (Condition B vs Condition A)
res <- results(dds)

# 4. Convert this object into a dataframe for easier visualisation

res <- as.data.frame(res)

# 5. Extract normalised counts for downstream machine learning
vsd <- vst(dds, blind = FALSE) # Variance stabilising transformation
normalized_counts <- assay(vsd)

```

## Step 3: Report DEGs (Table and Volcano Plot)

We define statistically significant DEGs as those with an adjusted p-value (FDR) \< 0.05 and an absolute log2 Fold Change \> 1.

```{r}
# Add -log10 adjusted p-value column
res$log10_p.adjust <- -log10(res$padj)

# Count significant DEGs
sig_degs <- res[which(res$padj < 0.05 & abs(res$log2FoldChange) > 1), ]
cat("Number of significant DEGs:", nrow(sig_degs), "\n")

# Show top 10
head(sig_degs[order(sig_degs$padj), ], 10)
```

Generate a volcano plot to visualise the results

```{r}
library(EnhancedVolcano)

EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'padj',
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10]~ 'adjusted p-value'),
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.0,
    labSize = 3.0,
    title = 'Cond B vs Cond A',
    subtitle = 'Differential expression')

```

## Step 4: Isolate Differentially Expressed Genes

To reduce noise and dimensionality for our machine learning algorithms, we will subset our normalized count matrix to only include the statistically significant DEGs identified above.

In order to make sure we don't lose too much information, let's relax a bit the thresholds for significance

```{r}
# Relaxed thresholds to keep enough genes for ML
relaxed_degs <- res[which(res$padj < 0.1 & abs(res$log2FoldChange) > 0.5), ]
cat("Number of DEGs (relaxed):", nrow(relaxed_degs), "\n")

# Subset normalised counts to DEG genes only
deg_counts <- normalized_counts[rownames(relaxed_degs), ]

# Transpose so samples are rows, genes are columns (needed for ML)
deg_matrix <- t(deg_counts)

# Condition labels
condition_labels <- colData$Condition

cat("Matrix dimensions:", nrow(deg_matrix), "samples x", ncol(deg_matrix), "genes\n")

```

## Step 5: Unsupervised Clustering

Let's see if the samples naturally group by their condition without explicitly telling the algorithm which sample belongs to which group. We will use Hierarchical Clustering and K-means.

## k-means

```{r}
set.seed(42)
km <- kmeans(deg_matrix, centers = 2, nstart = 25)
cat("K-means cluster assignments:\n")
table(km$cluster, condition_labels)
```

## hierarchical clustering

```{r}
# Calculate distance matrix and cluster
hc <- hclust(dist(deg_matrix), method = "complete")
# Cut into 2 groups
hc_clusters <- cutree(hc, k = 2)
cat("Hierarchical clustering assignments:\n")
table(hc_clusters, condition_labels)

```

Visualise the results (HINT: an heatmap would be perfect for this).

```{r}
library(pheatmap)

annotation <- data.frame(
  Condition = condition_labels,
  row.names = rownames(deg_matrix)
)

pheatmap(t(deg_matrix),
         annotation_col = annotation,
         show_rownames = FALSE,
         show_colnames = FALSE,
         scale = "row",
         clustering_method = "complete",
         main = "Hierarchical clustering of DEG expression")

```

## Step 6: Supervised Learning (LDA and SVM)

Next, we train classifiers to predict the condition of a sample based on its DEG expression profile. We will split the data into a training set (70%) and a testing set (30%).

```{r}
library(caret)
library(MASS)
library(e1071)

set.seed(42)
trainIndex <- createDataPartition(condition_labels, p = 0.7,
                                   list = FALSE, times = 1)
train_data   <- deg_matrix[trainIndex, ]
test_data    <- deg_matrix[-trainIndex, ]
train_labels <- condition_labels[trainIndex]
test_labels  <- condition_labels[-trainIndex]

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:", nrow(test_data), "\n")

```

```{r}
# Logistic Regression
train_df2 <- as.data.frame(train_data)
train_df2$Condition <- as.factor(train_labels)

model_LR <- glm(Condition ~ ., data = train_df2, family = 'binomial')

pred_LR_prob <- predict(model_LR, newdata = as.data.frame(test_data), type = 'response')
pred_LR <- ifelse(pred_LR_prob >= 0.5,
                  levels(as.factor(test_labels))[2],
                  levels(as.factor(test_labels))[1])

CM_LR <- confusionMatrix(data = as.factor(pred_LR), as.factor(test_labels))
CM_LR
```

Train LDA (Linear Discriminant Analysis)

```{r}
train_df <- as.data.frame(train_data)
train_df$Condition <- as.factor(train_labels)

model_LDA <- lda(Condition ~ ., data = train_df)

pred_LDA <- predict(model_LDA, newdata = as.data.frame(test_data))
CM_LDA <- confusionMatrix(data = as.factor(pred_LDA$class),
                           as.factor(test_labels))
CM_LDA
```

Train SVM (Support Vector Machine)

```{r}
model_SVM <- svm(x = train_data, y = as.factor(train_labels),
                  kernel = "linear")

pred_SVM <- predict(model_SVM, newdata = test_data)
CM_SVM <- confusionMatrix(data = as.factor(pred_SVM),
                           as.factor(test_labels))
CM_SVM
```

## Step 7: Compare Classifier Performance

Finally, we use confusion matrices to evaluate and compare how well LDA and SVM performed on the unseen test dataset.

```{r}
metrics <- data.frame(
  Classifier = c("Logistic Regression", "LDA", "SVM"),
  Accuracy = c(CM_LR$overall["Accuracy"],
               CM_LDA$overall["Accuracy"],
               CM_SVM$overall["Accuracy"]),
  Sensitivity = c(CM_LR$byClass["Sensitivity"],
                  CM_LDA$byClass["Sensitivity"],
                  CM_SVM$byClass["Sensitivity"]),
  Specificity = c(CM_LR$byClass["Specificity"],
                  CM_LDA$byClass["Specificity"],
                  CM_SVM$byClass["Specificity"]),
  Kappa = c(CM_LR$overall["Kappa"],
             CM_LDA$overall["Kappa"],
             CM_SVM$overall["Kappa"])
)
print(round(metrics[, -1], 4))
```

## Step 8: Is this enough?

is this enough to compare the classifiers? What if you use a k-fold cross-validation approach? Will you get similar results with different random splits?

```{r}
cv_df <- as.data.frame(deg_matrix)
cv_df$Condition <- as.factor(condition_labels)

set.seed(42)
folds_df <- fold(data = cv_df, k = 5, cat_col = "Condition")
fold_ids <- folds_df$.folds

out_LDA <- NULL
out_SVM <- NULL

for(k in 1:5){
  df_test  <- cv_df[fold_ids == k, ]
  df_train <- cv_df[fold_ids != k, ]
  test_lab  <- df_test$Condition
  train_lab <- df_train$Condition
  df_test$Condition  <- NULL
  df_train$Condition <- NULL

  # LDA
  m_lda <- lda(x = df_train, grouping = train_lab)
  p_lda <- predict(m_lda, newdata = df_test)$class
  cm_lda <- confusionMatrix(as.factor(p_lda), as.factor(test_lab))
  out_LDA <- rbind(out_LDA, cm_lda$overall)

  # SVM
  m_svm <- svm(x = as.matrix(df_train),
                y = as.factor(train_lab), kernel = "linear")
  p_svm <- predict(m_svm, newdata = as.matrix(df_test))
  cm_svm <- confusionMatrix(as.factor(p_svm), as.factor(test_lab))
  out_SVM <- rbind(out_SVM, cm_svm$overall)
}

cat("LDA 5-fold CV mean accuracy:", round(mean(out_LDA[,"Accuracy"]), 4), "\n")
cat("SVM 5-fold CV mean accuracy:", round(mean(out_SVM[,"Accuracy"]), 4), "\n")
cat("\nLDA per-fold accuracies:", round(out_LDA[,"Accuracy"], 4), "\n")
cat("SVM per-fold accuracies:", round(out_SVM[,"Accuracy"], 4), "\n")
```
